\section{Conclusion}

This paper presented Helios, a path-driven execution engine that accelerates EVM transaction processing while preserving gas-semantic correctness. On modern interpreters such as Revm, three barriers impede meaningful acceleration: the \textit{Compilation Limitation} where aggressive optimizations violate gas accounting, the \textit{Optimization Dilemma} where instrumentation costs exceed execution latency, and the \textit{Granularity Mismatch} where transaction-level caching suffers from combinatorial path explosion. Helios addresses these challenges through \textit{hybrid execution} that restricts optimization to static-cost instructions, \textit{asynchronous tracing} that decouples profiling from the critical path, and \textit{frame-level caching} that exploits Pareto-like call locality while resisting cache-flooding attacks. A guarded register-based interpreter executes the resulting SSA graphs, triggering native fallback upon path divergence to ensure correctness.

Helios advances the state of the art along four dimensions: \textit{safety} through hybrid execution and chunked gas accounting that maintain bit-level equivalence with canonical EVM semantics; \textit{efficiency} through asynchronous tracing and direct-access register files that remove instrumentation and stack manipulation from the critical path; \textit{robustness} through frequency-based filtering and bounded resource allocation that resist adversarial path explosion; and \textit{versatility} through a dual-mode architecture that serves both latency-sensitive validators and throughput-oriented archive nodes from a unified artifact set.

Evaluation on Ethereum mainnet confirms the effectiveness of this design: Helios achieves 6.60$\times$ median speedup in Replay Mode and 2.05$\times$ in Online Mode with 4.9\% storage overhead. Future work will explore selective JIT compilation for hot paths to further reduce interpretation overhead, frame-level fallback mechanisms to improve Online Mode coverage, and hybrid cache granularity that dynamically coalesces basic blocks into superblocks. By decoupling optimization from the critical path, Helios offers a foundation for scalable EVM infrastructure.
